% Easy Demographics Tables
% Toby Johnson <Toby.x.Johnson@gsk.com>
% `r format(Sys.time(), "%d %B %Y")`

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Easy Demographics Tables}
-->

```{r knitr-package, echo=FALSE}
suppressPackageStartupMessages(library(knitr))
```

# Making demographics tables easily and painlessly

A common component of both clinical and statistical genetics analyses
is the generation of "demographics tables".  From a statistical or
programming perspective, such tables are nothing more than summaries
of (typically several) ``demographic'' variables, summarized for one
or more groups of subjects.  A classic use case is summarising
baseline characteristics by treatment arm in a randomized clinical
study.  However, under the general definition used here, "demographics
tables" could summarize analysis endpoints, explanatory variables such
as genotypes, covariates, or indeed any other variable, by any set of
groups.  The groups may be disjoint or may be overlapping, for example
groups by genotype at a variant of interest, or one group for all
subjects in a clinical study and a second group for the subset of
subjects included in a genetic analysis.

The `demographics` functions in the gtx package aim to make
generation of demographics tables as easy as possible, given the
following assumptions:

* All analysis variables are already assembled into a single
    data frame, with one row per subject
* The variables in the data frame already have classes
    appropriate to their intended use in the analysis, e.g. factor,
    logical, numeric, or e.g. the `Surv` class for survival endpoint
* The output is to be optimized for *human*-readability, not for
    machine-readability, and hence a single cell in the output table
    may contain a mean (followed by a SD or range in parentheses),
    or a count (followed by a percentage in parentheses).

All the `demographics` functions output *character* arrays, which are
easily read in a fixed width font, and which may be easily inserted
into dynamically generated documents using e.g. `knitr::kable` or
`xtable::xtable`.

Under these assumtions, with analysis variables in a data frame `x`,
generating demographics by levels of the `TRTGRP` variable in `x` can be done as easily as:
```{r example1, eval=FALSE}
demographics(x, by = "TRTGRP")
```
For a more complicated example, generating demographics by two overlapping
groups that are defined on the fly, could be done quite easily using:
```{r example2, eval=FALSE}
demographics(x, by = list("All subjects" = rep(TRUE, nrow(x)),
                          Genotyped = x$ID %in% genotypes$ID))
```

For numerical summaries, different styles and decimal precisions can be specified in a straightforward and intuitive
way, both globally (here for all numeric variables) and for specific variables (here for AGE):
```{r example3, eval=FALSE}
demographics(x, by = list("All subjects" = rep(TRUE, nrow(x)),
                          Genotyped = x$ID %in% genotypes$ID),
             style = list(numeric = "Median (IQR)", AGE = "Median (Range)"),
             digits = list(numeric = 2, AGE = 0))
```

The following sections describe in more detail how to use the `demographics` functions.

## Specifying the demographic variables

Specifying the demographic variables to be summarized is easy: Put
them all in a dataframe and pass this as the first argument to
`demographics()`.  To summarize a restricted set of variables, the
data frame can be subsetted using base R functions `subset` or `[`
before passing to `demographics()`.  If a single variable is to be
summarized, it is generally better to pass it to `demographics()` as a
single column data frame (subset with `drop=FALSE`).

The behaviour of `demographics()` depends on the class(es) of its
arguments.  Currently, meaningful summaries are only produced for
classes `data.frame`, `logical`, `factor`, `numeric` and `Surv`.  In
particular, no meaningful summary is produced for arguments of
`character` class.  Therefore, variables in the input data frame must
be converted or coerced to appropriate classes.

An option `clinical.descriptors` can be used to provide a named
list of descriptive names to substitute for the actual
variable names.

An example of preparing suitable input, using the `aoex1` example
dataset:
```{r prepare}
suppressPackageStartupMessages(library(gtx))
options(clinical.usubjid = "USUBJID")
suppressPackageStartupMessages(library(survival))
data(aoex1)
aoex1 <- within(aoex1, {
  SEX <- factor(SEX) # convert to factor
  Male <- SEX == "Male" # logical
  OS <- Surv(SRVDY*12/365.25, SRVCFLCD) # survival, in months
  Genotyped <- !is.na(rs123456) # logical
})
options(clinical.descriptors = list(SEX = "Gender", OS = "Overall survival, Months",
                                    AGE = "Age, Years", bmi = "BMI, kg/m2"))
```

## Specifying the subgroups

The examples above illustrate the two principal ways that the `by`
argument can be used, either as a character vector or as a list of
logical vectors:

* If `by` is a character vector, it is interpreted as names of columns
    in `x`.  These columns must be either of logical or factor
    vectors.  Each logical vector is used to define exactly one group.
    Each factor vector is used to define one group per level of the
    factor.
* If `by` is a list of logical vectors, each is used to define exactly
    one group.  This second approach allows group membership to be
    determined without constructing additional columns in `x` and also
    allows finer control of the group titles.

# Demonstrations

The following demonstrations all use the `aoex1` dataset as modified above.
The `knitr::kable` function converts the character matrix produced by
`demographics()` into a markdown table that is rendered nicely to
appear in this vignette.

## Demographics by arm for one clinical study

In the first demonstration, we take a subset of rows in `aoex1` for
subjects in study AO000002 only, and a subset of columns for the
variables we want to summarize.  We define groups by `TRTGRP` which is
one of the factors in the data frame.  Note that because `TRTGRP` is
used to define groups it is automatically excluded from the variables
actually summarized.  This demonstration also illustrates a difference
between the summaries produced for factors (such as `SEX`) and for
logicals (such as `Male`).
```{r demo1, results="asis"}
kable(demographics(subset(aoex1, STUDYID == "AO000002",
                          select = c("SEX", "AGE", "Male", "bmi", "TRTGRP")),
                   by = "TRTGRP"))
```

## Changing the summary style for one variable

The summary style can be chosen by passing a named list as the
optional `style` argument, where the names of the list must match the
name of the variable for which the style is to be used (here `bmi`).
The style can be an arbitrary character string, and is converted to
produce the summary by simple text substitution.  Currently, the
following *case sensitive* patterns are substituted with numeric data:
"Mean", "Median", "SD", "Q1", "Q3", "Min", "Max", "IQR", "Range".
Note that the style string is also used as part of the row label.
Numerical precision can be controlled in a similar fashion by passing
a named list as the options `digits` argument.
```{r demo2, results="asis"}
kable(demographics(subset(aoex1, STUDYID == "AO000002",
                          select = c("SEX", "AGE", "bmi", "TRTGRP")),
                   by = "TRTGRP",
		   style = list(bmi = "Median (Range)"),
		   digits = list(bmi = 1)))
```

For tabulated count data (objects of class `factor`), the style string
substitutes *case sensitively* the patterns "N", "pc0", "pc1", and
"pc2" and "pc", which are respectively the count, percentages to 0, 1
and 2 decimals, and percentage to `digits` decimals.  The default
style is "N (pc0%)".  When the row label is constructed from the style
string, any "pc[012]*%*" patterns are changed to "%".  The style
commonly used in clinical statistics (without any percent sign in the
data columns) is like "N (pc1)".
```{r demo2bis, results="asis"}
kable(demographics(subset(aoex1, STUDYID == "AO000002",
                          select = c("SEX", "AGE", "bmi", "TRTGRP")),
                   by = "TRTGRP",
		   style = list(SEX = "N (pc1%)")))
```

## Changing the summary style for all variables

The `style` argument can also include elements named for classes of
variables, namely logical, factor, numeric and Surv.  If a variable
being sumamrized does not match `style` but does match by class, the
specified style will be used.  The `digits` argument works similarly.
Hence, in the following, the "Mean (IQR)" style and 0 decimal precision
will be used for all numeric variables except `bmi`, for which the
"Median (IQR)" style and 1 decimal precision will be used.
```{r demo3, results="asis"}
kable(demographics(subset(aoex1, STUDYID == "AO000002",
                          select = c("SEX", "AGE", "bmi", "TRTGRP")),
                   by = "TRTGRP",
		   style = list(bmi = "Median (IQR)", numeric = "Mean (IQR)"),
		   digits = list(bmi = 1, numeric = 0)))
```

## Defining more complex subgroups, by existing variables

Multiple variables in the data frame can be used to construct
subgroups, and they can be a mixture of factors (such as `STUDYID`)
and logicals (such as `Genotyped`).  Note that factors generate one
group per level in the factor, and logicals generate exactly one group
corresponding to subjects for which the logical is true.
```{r demo4, results="asis"}
kable(demographics(subset(aoex1,
                          select = c("bmi", "TRTGRP", "OS", "STUDYID", "Genotyped")),
	           by = c("STUDYID", "Genotyped")))
```

## Defining more complex subgroups, on the fly

Subgroups can also be constructed as a named list of logical vectors.
One advantage of this method is that the group names that appear in
the column names are constructed using the names of the list.  This
allows finer control than the "VARIABLE=value" column names generated
in the previous examples.  The names have to be quoted if they contain
spaces.  This demonstration includes a deliberately empty group, which
is automatically suppressed in the output (unless
`options(demographics.omit0groups = FALSE)` is used).
```{r demo5, results="asis"}
bygroups <- list("All subjects" = rep(TRUE, nrow(aoex1)),
                 "Study 1" = aoex1$STUDYID == "AO000001",
                 "Study 2" = aoex1$STUDYID == "AO000002",
                 "Men from Mars" = rep(FALSE, nrow(aoex1)))
kable(demographics(subset(aoex1, select = c("bmi", "TRTGRP")),
                   by = bygroups))
```

## No subgroups

If no subgroups are specified deliberately, or if subgroups are
improperly specified (names of variables not present in the first
argument, logicals of the wrong length, etc.), then a warning message
is issued and a failsafe group called "All subjects" is used.  This
demonstration also illustrates `demographics()` called on the whole
`aoex1` dataframe.
```{r demo6, results="asis"}
kable(demographics(aoex1, by = "MisTypEdNAme"))
```

## Survival endpoints

Variables of class `Surv` are handled in essentially a similar way to
numeric variables.  A style argument can be used, and the following
*case sensitive* patterns are substituted with numeric data: "Events",
"Median", "95% LCL", "95% UCL", "95% CI".  Note that some of these
patterns contain spaces.  This demonstration also illustrates that
because styles work by text substitution, it is easy to change the
style to use square brackets instead of parantheses.
```{r demo7, results="asis"}
kable(demographics(subset(aoex1,
                          select = c("bmi", "TRTGRP", "OS", "STUDYID", "Genotyped")),
                   by = bygroups,
		   style = list(Surv = "Median [95% CI]", numeric = "Mean [Range]")))
```

# Internal workings and further customization

`demographics` is a generic function and uses S3 method dispatching.
All the examples and demonstrations above use the
`demographics.data.frame()` method.  This method does some checking
and conversion of the `by` argument, finds the most relevant element
in the optional `style` argument, and then calls the generic
`demographics` for each variable in the data frame in turn.
Therefore, the user can in principle define new classes and write
`demographics` methods for those classes.

The `demographics` functions have some behaviours that are controlled
by options that can be set by the user.  Currently, these control
whether rows where all groups have zero count are omitted
(demographics.omit0rows, default TRUE), and whether groups with no
subjects are omitted (demographics.omit0groups, default TRUE).



In the future it is planned to provide automatic collapsing of levels
for factors with many levels.

# Why use the `demographics` functions?

Some of what the `demographics` functions do is provided by existing
functions such as `by`, `aggregate` and `summary.data.frame`.
However, none of these functions can do in one or two lines what the
`demographics` functions do, especially with regard to producing near
publication ready output.  In particular, the `demographics` functions
allow fine user control of how numerical data are summarised, and can
summarise survival endpoints in the manner widely used in the clinical
literature.

No attempt has currently been made to accomodate the finer aspects of
publication readiness.  Typically, some fine adjustment of spacing
will be needed, but this adjustment will depend on factors such as the
font used in the final publication, and therefore is difficult to
implement within the R package.

# About this document

This document is written in [R
Markdown](https://support.rstudio.com/hc/en-us/articles/200488468-R-Markdown),
using the [Pandoc extended and revised version of
Markdown](http://johnmacfarlane.net/pandoc/README.html#pandocs-markdown).
To make persistent edits you *need* to edit the source (.Rmd) file, not
any other files (such as the .md, .html or .docx file).  

To compile this document as a .docx (document), run the following commands in R:

    library(knitr)
    knit("demographics.Rmd")`
    pandoc("demographics.md", format = "docx")

To compile this document as a .html (web page), run the following commands in R:

    library(knitr)
    knit2html("demographics.Rmd")

This document was complied on `r date()` using `r R.version.string`

